---
title: "Panel Data Analysis"
author: "Kihwan Yu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---




```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# is that good?22wwwwwww

# test test

# godgodgggggggg

# Iwant to change that

### 0.1. how to do use Git?

     -  How to pull?
     -  How to push?
          1. make commitment
          2. push.
          Note that we must to do make commitment first. Why? we need to write down the descriptions about the version.
     
### What am I supposed to do if I meet these errors?

     1. 
     
     
 
## 0. Introduction
     

나의 지도 교수님이 추천해주신 책의 내용을 정리하면서, 공부를 시작해 볼 예정. 이 material들의 목적은 다음과 같다.

1.  나의 공부

    1.1. 이론을 공부한다.

    1.2. 동시에 형실에 응용할 수 있는 감을 익힌다.

2.  사람들에게 공부한것을 공유하면서 뿌듯함을 느낀다.

3.  나중에 학생을 가르칠 일이 있을 때를 대비 한다.

4.  지금 정리해두면 잡마켓에 도움이 되지 않을까?

5.  그럼 시작 !

6.  학기중에도 작성해야지

7.  why??????

8.  그래 바껴야지

9.  plzplzplzplz

10 attemp

10 whay???

1

## 1. Panel Data - 개요

### What is Panel Data?

패널 데이터는 같은 대상들에 대해 일정 기간동안 반복해서 관찰한 데이터를 의미한다.

### What is advantage of Panel Data?

1.  omitted variables로 인한 bias들은 통제할 수 있게 된다.

### What is disvantage of Panel Data?

1.  High cost to maintain data

2.  If run standard OLS using panel data, then we may have biased estimates. see followings

## 2. settings - the situation we are looking through

### 2.0. What data do we need to have?

Let we have [balanced panel data]{.underline}.

-   What is panel data and balanced panel data?

    -   Panel data are the data that observed units repeatedly over time
    -   Balanced panel data are the data that observed every i in every t, meaning that no missing in the time period

### 2.1. notation settings

LHS var for individual i at time t is that $$y_{it}, \ \ \  i = 1,...N, and t = 1,...,T.$$

```{r}
set.seed(6262)

#set T = 3 and N = 4 to demonstrate example

#y11, ..., y14
     #y1t, t = 1,...4
     y_i1_t1 <- rnorm(1)
     y_i1_t2 <- rnorm(1)
     y_i1_t3 <- rnorm(1)
     
     #y2t, t = 1,...4
     y_i2_t1 <- rnorm(1)
     y_i2_t2 <- rnorm(1)
     y_i2_t3 <- rnorm(1)
     
     #y3t, t = 1,...4
     y_i3_t1 <- rnorm(1);  y_i3_t2 <- rnorm(1);  y_i3_t3 <- rnorm(1);
     y_i4_t1 <- rnorm(1);  y_i4_t2 <- rnorm(1);  y_i4_t3 <- rnorm(1);
```

Using vector notation, we can denote the LHS variable such that

$${\bar{Y}}_i=\left(y_{i1},\ \ldots,\ y_{iT}\right)^\prime,$$

```{r}
#yi, i = 1,...9
     #y1
     y_i1 <- t(cbind(y_i1_t1, y_i1_t2, y_i1_t3))
     y_i2 <- t(cbind(y_i2_t1, y_i2_t2, y_i2_t3))
     y_i3 <- t(cbind(y_i3_t1, y_i3_t2, y_i3_t3))
     y_i4 <- t(cbind(y_i4_t1, y_i4_t2, y_i4_t3))
               
     print(y_i1)
     print(y_i4)
```

To use matrix notation, we can denote this in convenience such that

$${\bar{Y}}=\left(y_{i},\ \ldots,\ y_{i}\right)^\prime,$$

```{r}
#yi, i = 1,...9
     #y1
     y <- rbind(y_i1, y_i2, y_i3, y_i4)
     print(y)
```

K-th RHS variable for individual i at time t is that

$$x_{itk}, \ \ \  i = 1,...N, \  t = 1,...,T, and \ k = 1,..., K.$$

```{R}
#set T = 3, N = 4, K = 2 to demonstrate example
#x_11k for k = 1,2
     #x11k, k = 1, 2
     x_i1_t1_k1 <- rnorm(1)
     x_i1_t1_k2 <- rnorm(1)
     
     #x_1tk,, t = 1,..., 4 k = 1, 2
     x_i1_t2_k1 <- rnorm(1)
     x_i1_t2_k2 <- rnorm(1)
     x_i1_t3_k1 <- rnorm(1)
     x_i1_t3_k2 <- rnorm(1)

     #x_itk, i=1,...,9 t = 1,..., 4 k = 1, 2
     x_i2_t1_k1 <- rnorm(1); x_i2_t1_k2 <- rnorm(1); x_i2_t2_k1 <- rnorm(1); x_i2_t2_k2 <- rnorm(1); x_i2_t3_k1 <- rnorm(1); x_i2_t3_k2 <- rnorm(1); 
     x_i3_t1_k1 <- rnorm(1); x_i3_t1_k2 <- rnorm(1); x_i3_t2_k1 <- rnorm(1); x_i3_t2_k2 <- rnorm(1); x_i3_t3_k1 <- rnorm(1); x_i3_t3_k2 <- rnorm(1); 
     x_i4_t1_k1 <- rnorm(1); x_i4_t1_k2 <- rnorm(1); x_i4_t2_k1 <- rnorm(1); x_i4_t2_k2 <- rnorm(1); x_i4_t3_k1 <- rnorm(1); x_i4_t3_k2 <- rnorm(1);
```

Using vector notation, we have $${\bar{x}}_{it}=\left(x_{it1},\ \ldots,\ x_{itK}\right)^\prime,$$

```{R}
#x_it
     #x_1t, t = 1,..., 4.
     x_i1_t1 <- (cbind(x_i1_t1_k1, x_i1_t1_k2))
     x_i1_t2 <- (cbind(x_i1_t2_k1, x_i1_t2_k2))
     x_i1_t3 <- (cbind(x_i1_t3_k1, x_i1_t3_k2))
     
     #x_it, i=1,...,9, t = 1,..., 4.
     x_i2_t1 <- (cbind(x_i2_t1_k1, x_i2_t1_k2));  x_i2_t2 <- (cbind(x_i2_t2_k1, x_i2_t2_k2));  x_i2_t3 <- (cbind(x_i2_t3_k1, x_i2_t3_k2));  
     x_i3_t1 <- (cbind(x_i3_t1_k1, x_i3_t1_k2));  x_i3_t2 <- (cbind(x_i3_t2_k1, x_i3_t2_k2));  x_i3_t3 <- (cbind(x_i3_t3_k1, x_i3_t3_k2));  
     x_i4_t1 <- (cbind(x_i4_t1_k1, x_i4_t1_k2));  x_i4_t2 <- (cbind(x_i4_t2_k1, x_i4_t2_k2));  x_i4_t3 <- (cbind(x_i4_t3_k1, x_i4_t3_k2));  

     print(x_i1_t1)
     print(x_i4_t3)


```

Using matrix notation, we have $${\bar{X}}_i=\left(\begin{matrix}x_{i11}&\cdots&x_{i1T}\\\vdots&\ddots&\vdots\\x_{iK1}&\cdots&x_{iKT}\\\end{matrix}\right)^\prime,$$

```{r}
#xi
     #x1
     x_i1 <- rbind((x_i1_t1), (x_i1_t2), (x_i1_t3))

     #xi, i=1,...9
     x_i2 <- rbind((x_i2_t1), (x_i2_t2), (x_i2_t3))
     x_i3 <- rbind((x_i3_t1), (x_i3_t2), (x_i3_t3))
     x_i4 <- rbind((x_i4_t1), (x_i4_t2), (x_i4_t3))


     print(x_i1)
     print(x_i4)
```

Using matrix notation for once more, we have $${\bar{X}}=\left(x_{1},\ \ldots,\ x_{N}\right)^\prime,$$

```{r}
#x
     x <- rbind(x_i1, x_i2, x_i3, x_i4)

     print(x)
```

also, we have $$\bar{\beta}=\left(\beta_1,\ \ldots,\ \beta_k\right)$$

$${\bar{\varepsilon}}_i=\left(\varepsilon_{i1},\ \ldots,\ \varepsilon_{iT}\right)^\prime.$$

### 2.2. Specification

$${\bar{Y}}_i=\ {\bar{X}}_i^\prime\bar{\beta}+{\bar{\varepsilon}}_i,\ \ \ \ i=1,\ \ldots,\ n, $$

$$where\ {\bar{\varepsilon}}_i={\bar{a}}_i+{\bar{u}}_i,\ \ \ {\bar{a}}_i=\left(\begin{matrix}a_i&\ldots&a_i\\\end{matrix}\right),\ and\ {\bar{u}}_i=\left(\begin{matrix}u_{i1}&\ldots&u_{it}\\\end{matrix}\right)\ $$

In this specification, We called ${\bar{a}}_i$ to time-invariant error term, and ${\bar{u}}_i$ to time-variant error term.

## 3. Is this identified?

YES, This equation could be identified.

$${\bar{Y}}_i=\ {\bar{X}}_i^\prime\bar{\beta}+{\bar{\varepsilon}}_i,\ \ \ \ i=1,\ \ldots,\ n$$

$${\bar{X}}_i{\bar{Y}}_i=\ {{\bar{X}}_i\bar{X}}_i^\prime\bar{\beta}+{{\bar{X}}_i\bar{\varepsilon}}_i$$

$$E[{\bar{X}}_i{\bar{Y}}_i]= E[XiXi']β+E[X_iε_i]$$

$$\bar{\beta}=E\left[{{\bar{X}}_i\bar{X}}_i^\prime\right]^{-1}E\left[{\bar{X}}_i{\bar{Y}}_i\right]-E\left[{{\bar{X}}_i\bar{X}}_i^\prime\right]^{-1}E\left[{{\bar{X}}_i\bar{\varepsilon}}_i\right].$$

$$\bar{\beta}=E\left[{{\bar{X}}_i\bar{X}}_i^\prime\right]^{-1}E\left[{\bar{X}}_i{\bar{Y}}_i\right] -E[{\bar{X}}_ia_i] - 0$$

Thus, this equation is identified if $E\left[{{\bar{X}}_i\bar{\varepsilon}}_i\right]$ is equal to zero, meaning that we need to have $E[{\bar{X}}_i a_i] = 0$ and $E[{\bar{X}}_i u_i] = 0$.

## 4. Estimation

### 4.1. Pooled ols case

#### 4.1.1. what is the pooled ols?

-   a standard ols estimator running with panel data

#### 4.1.2. pooled ols estimator

the corresponding pooled ols estimator is that

$${\widehat{\bar{\beta}}}_{ols}=\left(\frac{1}{n}\sum_{i}^{n}{{\bar{X}}^\prime_i{\bar{X}}_i}\right)^{-1}\left(\frac{1}{n}\sum_{i}^{n}{{\bar{X}}^\prime_i{\bar{Y}}_i}\right)$$

```{r}
#checking dimension
dim(x_i1)
dim(y_i1)
dim(t(x_i1)%*%(y_i1))
dim(t(x_i1)%*%(x_i1))

# run using vector notation
     # making first term of RHS
     ave_Xi_Xi <- (1/4) * ( t(x_i1)%*%(x_i1) + t(x_i2)%*%(x_i2) + t(x_i3)%*%(x_i3) + t(x_i4)%*%(x_i4) )
     inv_ave_Xi_Xi <- solve(ave_Xi_Xi)
     
     # making second ferm of RHS
     ave_Xi_Yi <- (1/4) * ( t(x_i1)%*%(y_i1) + t(x_i2)%*%(y_i2) + t(x_i3)%*%(y_i3) + t(x_i4)%*%(y_i4) )   

     #calculatr beta
     Beta_pooled <- inv_ave_Xi_Xi %*% ave_Xi_Yi

     print(Beta_pooled)

# run using R function
     ols_res <- lm(y ~ -1 +x[,1] + x[,2]) # -1 mean no constant
     ols_res

```

#### 4.1.3. Is this a consistent estimator?

NO, unless $E[{\bar{X}}_i a_i] = 0$

$${\widehat{\bar{\beta}}}_{ols}=\left(\frac{1}{n}\sum_{i}^{n}{{\bar{X}}_i{\bar{X}}_i^\prime}\right)^{-1}\left(\frac{1}{n}\sum_{i}^{n}{{\bar{X}}_i{\bar{Y}}_i}\right)$$

$${\widehat{\bar{\beta}}}_{ols}=\left(\frac{1}{n}\sum_{i}^{n}{{\bar{X}}_i{\bar{X}}_i^\prime}\right)^{-1}\left(\frac{1}{n}\sum_{i}^{n}{{\bar{X}}_i({\bar{X}}_i^\prime\bar{\beta}+{\bar{a}}_i+{\bar{u}}_i)}\right)$$

$${\widehat{\bar{\beta}}}_{ols}={\bar{\beta}+\left(\frac{1}{n}\sum_{i}^{n}{{\bar{X}}_i{\bar{X}}_i^\prime}\right)}^{-1}\left(\frac{1}{n}\sum_{i}^{n}{{\bar{X}}_i{\bar{a}}_i}\right){+\left(\frac{1}{n}\sum_{i}^{n}{{\bar{X}}_i{\bar{X}}_i^\prime}\right)}^{-1}\left(\frac{1}{n}\sum_{i}^{n}{{\bar{X}}_i{\bar{u}}_i}\right)$$

$$plim\left({\widehat{\bar{\beta}}}_{ols}\right)\ =\ \bar{\beta}+E\left[{\bar{X}}_i{\bar{X}}_i^\prime\right]^{-1}E\left[{\bar{X}}_i{\bar{a}}_i\right]+0$$

### 4.2. Fixed (within) effects model

Intuition: transform data to eliminate $a_i$ form error term. It only works if ① $a_i$ is time invariant, ② ${\bar{X}}_i$ is time varying.

#### 4.2.1. New Notation and Specification

Let use a new notation,

$${\ddot{y}}_{it}=y_{it}-\frac{1}{T}\sum_{t=1}^{T}y_{it},$$

$${\ddot{X}}_{it}=X_{it}-\frac{1}{T}\sum_{t=1}^{T}X_{it}, and$$

$${\ddot{u}}_{it}=u_{it}-\frac{1}{T}\sum_{t=1}^{T}u_{it}$$

Then, we have a new specification such that $${\ddot{y}}_{it}=\ {\ddot{X}}_i^\prime\bar{\beta}+{\ddot{u}}_i\ $$

#### 4.2.2. Is this specification same with the normal ols specification?

Yes,

$${\bar{Y}}_i=\ {\bar{X}}_i^\prime\bar{\beta}+{\bar{\varepsilon}}_i, $$

$${\bar{Y}}_i-\frac{1}{T}\sum_{t=1}^{T}y_{it}=\ {\bar{X}}_i^\prime\bar{\beta}+{\bar{\varepsilon}}_i\ -\frac{1}{T}\sum_{t=1}^{T}y_{it}$$ $${\ddot{y}}_{it}=\ {\bar{X}}_i^\prime\bar{\beta}+{\bar{\varepsilon}}_i\ -\frac{1}{T}\sum_{t=1}^{T}\left({\bar{X}}_i^\prime\bar{\beta}+{\bar{\varepsilon}}_i\right)$$

$${\ddot{y}}_{it}=\ {\bar{X}}_i^\prime\bar{\beta}+{\bar{a}}_i+{\bar{u}}_i\ -\frac{1}{T}\sum_{t=1}^{T}\left({\bar{X}}_i^\prime\bar{\beta}+{\bar{a}}_i+{\bar{u}}_{it}\right)$$

$${\ddot{y}}_{it}=\ {\bar{X}}_i^\prime\bar{\beta}+{\bar{u}}_i\ -\frac{1}{T}\sum_{t=1}^{T}{{\bar{X}}_i^\prime\bar{\beta}}-\frac{1}{T}\sum_{t=1}^{T}{\bar{u}}_{it}$$

$${\ddot{y}}_{it}=\ {\ddot{X}}_i^\prime\bar{\beta}+{\ddot{u}}_i\ $$

#### 4.2.3. then, what is FEs estimator?

$${\widehat{\bar{\beta}}}_{FE}=\left(\frac{1}{n}\sum_{i=1}^{n}{\ddot{\bar{X}}}_i{\ddot{\bar{X}}}_i^\prime\right)^{-1}\left(\frac{1}{n}\sum_{i=1}^{n}{\ddot{\bar{X}}}_i{\ddot{\bar{Y}}\ }_i\right)$$

#### 4.2.3. Is this a consistent estimator ?

YES,

$${\widehat{\bar{\beta}}}_{FE}=\left(\frac{1}{n}\sum_{i=1}^{n}{\ddot{\bar{X}}}_i{\ddot{\bar{X}}}_i^\prime\right)^{-1}\left(\frac{1}{n}\sum_{i=1}^{n}{\ddot{\bar{X}}}_i{\ddot{\bar{Y}}\ }_i\right)\ $$

$${\widehat{\bar{\beta}}}_{FE}=\left(\frac{1}{n}\sum_{i=1}^{n}{\ddot{\bar{X}}}_i{\ddot{\bar{X}}}_i^\prime\right)^{-1}\left(\frac{1}{n}\sum_{i=1}^{n}{\ddot{\bar{X}}}_i({\ddot{X}}_i^\prime\bar{\beta}+{\ddot{u}}_i)\right)$$

$${\widehat{\bar{\beta}}}_{FE}=\bar{\beta}+\left(\frac{1}{n}\sum_{i=1}^{n}{\ddot{\bar{X}}}_i{\ddot{\bar{X}}}_i^\prime\right)^{-1}\left(\frac{1}{n}\sum_{i=1}^{n}{\ddot{\bar{X}}}_i{\ddot{u}}_i\right)$$

$$plim\left({\widehat{\bar{\beta}}}_{FE}\right)=\ \bar{\beta}$$

### 4.3. between effects model

## 5. example with code

```{r}
library(tidyverse)
library(haven)
library(estimatr)
library(plm)


read_data <- function(df)
{
  full_path <- paste("https://github.com/scunning1975/mixtape/raw/master/", 
                     df, sep = "")
  df <- read_dta(full_path)
  return(df)
}

sasp <- read_data("sasp_panel.dta")

#-- Delete all NA
sasp <- na.omit(sasp)

#-- order by id and session 
sasp <- sasp %>% 
  arrange(id, session)

#Balance Data
balanced_sasp <- make.pbalanced(sasp, 
                                balance.type = "shared.individuals")

#Demean Data
balanced_sasp <- balanced_sasp %>% mutate( 
  demean_lnw = lnw - ave(lnw, id),
  demean_age = age - ave(age, id),
  demean_asq = asq - ave(asq, id),
  demean_bmi = bmi - ave(bmi, id),
  demean_hispanic = hispanic - ave(hispanic, id),
  demean_black = black - ave(black, id),
  demean_other = other - ave(other, id),
  demean_asian = asian - ave(asian, id),
  demean_schooling = schooling - ave(schooling, id),
  demean_cohab = cohab - ave(cohab, id),
  demean_married = married - ave(married, id),
  demean_divorced = divorced - ave(divorced, id),
  demean_separated = separated - ave(separated, id),
  demean_age_cl = age_cl - ave(age_cl, id),
  demean_unsafe = unsafe - ave(unsafe, id),
  demean_llength = llength - ave(llength, id),
  demean_reg = reg - ave(reg, id),
  demean_asq_cl = asq_cl - ave(asq_cl, id),
  demean_appearance_cl = appearance_cl - ave(appearance_cl, id),
  demean_provider_second = provider_second - ave(provider_second, id),
  demean_asian_cl = asian_cl - ave(asian_cl, id),
  demean_black_cl = black_cl - ave(black_cl, id),
  demean_hispanic_cl = hispanic_cl - ave(hispanic_cl, id),
  demean_othrace_cl = othrace_cl - ave(othrace_cl, id),
  demean_hot = hot - ave(hot, id),
  demean_massage_cl = massage_cl - ave(massage_cl, id)
  )

#-- POLS
ols <- lm_robust(lnw ~ age + asq + bmi + hispanic + black + other + asian + schooling + cohab + married + divorced + separated + 
           age_cl + unsafe + llength + reg + asq_cl + appearance_cl + provider_second + asian_cl + black_cl + hispanic_cl + 
           othrace_cl + hot + massage_cl, data = balanced_sasp)
summary(ols)


#-- FE
formula <- as.formula("lnw ~ age + asq + bmi + hispanic + black + other + asian + schooling + 
                      cohab + married + divorced + separated + 
                      age_cl + unsafe + llength + reg + asq_cl + appearance_cl + 
                      provider_second + asian_cl + black_cl + hispanic_cl + 
                      othrace_cl + hot + massage_cl")

model_fe <- lm_robust(formula = formula,
                  data = balanced_sasp, 
                  fixed_effect = ~id, 
                  se_type = "stata")

summary(model_fe)

#-- Demean OLS
dm_formula <- as.formula("demean_lnw ~ demean_age + demean_asq + demean_bmi + 
                demean_hispanic + demean_black + demean_other +
                demean_asian + demean_schooling + demean_cohab + 
                demean_married + demean_divorced + demean_separated +
                demean_age_cl + demean_unsafe + demean_llength + demean_reg + 
                demean_asq_cl + demean_appearance_cl + 
                demean_provider_second + demean_asian_cl + demean_black_cl + 
                demean_hispanic_cl + demean_othrace_cl +
                demean_hot + demean_massage_cl")

ols_demean <- lm_robust(formula = dm_formula, 
                data = balanced_sasp, clusters = id,
                se_type = "stata")

summary(ols_demean)

```

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
